Processing genome scale tabular data with wormtable

### 2013-08-21 docs/paper/paper.tex
RN is done his edits

Wormtable is currently limited to supporting Python only, and 
another important aspect of future development is to 
develop a C library along with bindings for other popular
languages such as Perl, R and PHP. 
Wormtable does not support interval search, and so it is not
straightforward to find, for example, all rows overlapping 
a given genomic region in GTF files. 
This problem has been 
solved several times, however, and we aim to adapt existing 
techniques~\cite{li11,kzbhk10} and incorporate them into 
wormtable.  Compression of data can
result in poor performance, but it is often necessary when 
volumes of data are very large. Thus, we plan on introducing 
optional compression of the data file in wormtable in 
a future release. Beyond these additions, it is difficult
predict the precise direction of future development since this depends 
on feedback from the community.  Wormtable is an open and 
collaborative project actively seeking feedback and 
contributors.

%%%%%%%%%%%%%%%%%%%%%%
\section*{Conclusions}
%\emph{This should state clearly the main conclusions of the article and give a clear 
%explanation of the importance and relevance of the software.}
%
The volume of data being produced in bioinformatics is constantly 
increasing, but the tools available to end users to 
process data are still mostly based on parsing plain text. This 
approach is very inefficient, and leads to several undesirable outcomes.
Firstly, and most obviously, a researcher's productivity is inevitably
reduced by being forced to wait for several hours for the result of a 
simple calculation. Without flexible indexing, working with 
a subset of a data file usually requires the creation of 
another file consisting of the subset in question, requiring 
extra storage and maintenance.
There are also some less obvious consequences:
code quality is reduced, since testing over the entire dataset is 
infeasible, and it is much less likely that the effects of 
changing arbitrary quality thresholds (for example) will be 
systematically examined.

The classical approach to solving problems of this type 
is to use a relational database, which provide sophisticated 
data management techniques. Relational databases are unsuitable 
for storing static datasets, however, as they are complex to 
use and incur many unnecessary overheads.
Wormtable provides the most important features of database
technologies (packed binary storage of values; random access 
to rows; general purpose indexing) without their complexities 
and overheads. Wormtable is also more flexible than 
relational databases, supporting, for example, columns 
containing a variable number of integers. Finally, wormtable 
is far more adaptable than a relational database.
Any widely used database system today is a hugely complex 
piece of software, and adding necessary features would be 
very difficult. Adding new features to wormtable, on the 
other hand, is straightforward since it is far simpler and 
does not need to be compatible with the relational model
and decades worth of existing software.

The most important aspect of wormtable is its efficiency
and ease of use for end users, and we illustrated these
points using some examples of VCF data.
After converting a file in VCF format to 
wormtable using \texttt{vcf2wt} a user can process the 
data very efficiently using Python. Accessing data from 
rows in wormtable is many times faster than is possible 
by parsing rows encoded as text.
This is also a very convenient 
way to access VCF data, since individual columns are already 
parsed and all that is needed is the name of the column of 
interest. To access regions of the genome efficiently
the user simply needs to create an index on the CHROM and POS columns
using the  \texttt{wtadmin add} command, and can then efficiently
seek to arbitrary locations. Wormtable is not 
limited to VCF, but can store
any form of fixed tabular data. We also provide a conversion 
tool for the GTF format, and several others are planned.

>>>>

Wormtable is currently limited to supporting Python, and 
another important aspect of future development is to 
develop a C library along with bindings for other popular
languages such as Perl, R and PHP. 
Wormtable does not support interval search, and so it is not
straightforward to find, for example, all rows overlapping 
a given genomic region in GTF files. 
This problem has been 
solved several times, however, and we aim to adapt existing 
techniques~\cite{li11,kzbhk10} and incorporate them into 
wormtable.  Compression of data can
result in poor performance, but it is often necessary when 
volumes of data are very large. Thus, we plan on introducing 
optional compression of the data file in wormtable in 
a future release. Beyond these additions, it is difficult
predict the precise direction of future development since this depends 
on feedback from the community.  Wormtable is an open and 
collaborative project actively seeking feedback and 
contributors.

%%%%%%%%%%%%%%%%%%%%%%
\section*{Conclusions}
%\emph{This should state clearly the main conclusions of the article and give a clear 
%explanation of the importance and relevance of the software.}
%
The volume of data being produced in biological research is growing 
rapidly, but the tools available to end users to 
process data are still mostly based on parsing plain text. This 
approach is very inefficient, and leads to several undesirable outcomes.
Firstly, and most obviously, a researcher's productivity is inevitably
constrained while waiting several hours for the result of a 
simple calculation. Without flexible indexing, working with 
a subset of a data file usually requires the creation of 
another file consisting of the subset in question, requiring 
extra storage and maintenance.
Additionally, code quality is reduced, since testing over 
the entire dataset is 
infeasible and it is less likely that the effects of 
changing arbitrary analysis parameters will be 
systematically examined.

The classical approach to solving problems of this type 
is to use a relational database, which provide sophisticated 
data management techniques. However, relational databases are unsuitable 
for storing static datasets as they are complex to 
use and incur many unnecessary overheads.
Wormtable provides the most important features of database
technologies (packed binary storage of values; random access 
to rows; general purpose indexing) without additional complexities 
and overheads. Wormtable is also more flexible than 
relational databases, supporting, for example, columns 
containing a variable number of integers. Finally, wormtable 
is far more adaptable than a relational database.
All widely used database systems are complex 
and adding the necessary features would be 
very difficult. Adding new features to wormtable, on the 
other hand, is straightforward because it is far simpler and 
does not need to be compatible with the relational model
and decades worth of existing software.

The most important aspect of wormtable is its efficiency
and ease of use for end users, and we illustrated these
points using some examples of VCF data.
After converting a file in VCF format to 
wormtable using \texttt{vcf2wt} a user can process the 
data very efficiently using Python. Accessing data from 
rows in wormtable is many times faster than is possible 
by parsing rows encoded as text.
This is also a very convenient 
way to access VCF data, since individual columns are already 
parsed and all that is needed is the name of the column of 
interest. To access regions of the genome efficiently
the user simply needs to create an index on the chromosome 
(CHROM) and position (POS) columns
using the  \texttt{wtadmin add} command. Wormtable is not 
limited to VCF, but can store 
any form of fixed tabular data. We provide a conversion 
tool for the GTF format, and several others are planned or 
could be contributed by users.
